A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Updates from Overleaf

The precise genomic location of a recombination breakpoint is 
in general impossible to determine, but we can place 
bounds by inspecting sequences of the recombinant and its parents.
The breakpoint interval of a recombinant is defined by
two positions: a left position where the recombinant
matches the left parent sequence but not the right, and
a right position where the recombinant matches the right
parent sequence but not the left. 
Genealogical methods such as sc2ts and UShER+RIPPLES
use inferred parental sequences and are therefore 
in principle maximally precise.
Other methods are limited in precision by the various proxies 
used to represent distinct parental \emph{lineages}~\cite{%
Varabyou2021rapid,Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}.
% How well the time and position of a recombination event
% can be inferred thus depends on the divergence between the
% parent sequences (measured as time to their MRCA).

The distribution of breakpoint intervals of
the QC-passing recombination events in the sc2ts ARG
shows an overall enrichment of breakpoints
towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with the patterns seen in previous
studies~\cite{Bobay2020recombination,Turakhia2022,
Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
making recombination events more likely to be detectable, and 
allowing breakpoint intervals to be estimated more precisely.
We found the median breakpoint interval length to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably larger (up to $\sim$54\% of the genome).
As expected from theory (STAR Methods), we found a negative relationship between
the length of a breakpoint interval and
divergence between recombinant parents
(Figure~\ref{fig:recombinant_intervals}D).
Breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

The time interval of a recombination node in the ARG is bounded by
the age of the youngest parent and
the age of the oldest child of the recombination node.
We found that for $\sim$75\% of the QC-passing recombination events,
the time interval size is less than $\sim$2 months
(Figure~\ref{fig:recombinant_intervals}D);
% Jackson samples in <= 7 day range:
% Group B ERR5058070 (5 day), singleton ERR5054123 (7 days)
for example, the time intervals of
the events associated with the Jackson et al.\ samples
span 5 to 48 days.
% Q3: 60 days, max: 259 days
Longer time intervals imply that a parental or a recombinant lineage
might have been unsampled for an extended period of time.

>>>>

The location of a recombination breakpoint is generally
impossible to pinpoint exactly, and instead an interval
containing the breakpoint can be inferred by
examining the parental and recombinant sequences.
Some methods~\cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
and approaches (e.g., Pango designation) infer this
as the genomic distance between lineage-associated
mutations that differentiate the two parental \emph{lineages}.
Similarly, methods based on probabilistic representations of
mutational profiles \cite{Varabyou2021rapid}
can only be as precise as examination of the two
parent lineages allows.
However, genealogical methods (such as sc2ts and RIPPLES) are
based on identifying exact parental \emph{sequences},
so are in principle maximally precise.
The breakpoint interval of a recombinant is bounded by
two positions: a left position where the recombinant
matches the left parent sequence but not the right, and
a right position where the recombinant matches the right
parent sequence but not the left. 
How well the time and position of a recombination event
can be inferred thus depends on the divergence between the
parent sequences (measured as time to their MRCA).

The distribution of breakpoint intervals of
the QC-passing recombination events
shows an overall enrichment of breakpoints
towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with the patterns seen in previous
studies~\cite{Bobay2020recombination,Turakhia2022,
Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
making recombination events more likely to be detectable, and 
allowing breakpoint intervals to be estimated more precisely.
We found the median breakpoint interval length to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably larger (up to $\sim$54\% of the genome).
As expected, we found a negative relationship between
the length of a breakpoint interval and
divergence between recombinant parents
(Figure~\ref{fig:recombinant_intervals}D),
which follows theoretical expectation (STAR Methods).
Breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

The time interval of a recombination node in the ARG is bounded by
the age of the youngest parent and
the age of the oldest child of the recombination node
(either given as collection dates or estimated using tsdate).
We found that for $\sim$75\% of the QC-passing recombination events,
the time interval size is less than $\sim$2 months
(Figure~\ref{fig:recombinant_intervals}D);
% Jackson samples in <= 7 day range:
% Group B ERR5058070 (5 day), singleton ERR5054123 (7 days)
for example, the time intervals of
the events associated with the Jackson et al.\ samples
span 5 to 48 days.
% Q3: 60 days, max: 259 days
Longer time intervals imply that a parental or a recombinant lineage
might have been unsampled for an extended period of time.
